MR-PFU: A suite of R packages for energy conversion chain analysis

### 2023-07-24 paper.Rmd
Added information about more packages.

Several decisions influenced the design and APIs for the packages.
The decision involves data format. 
We selected the Physical Supply-Use Table (PSUT) framework,
a matrix approach to describing energy flows from 
extraction to
processing stages and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018].
The PSUT framework succinctly describes the flow of energy carriers
("products" in PSUT terminology) 
among energy conversion machines 
("industries" in PSUT terminology)
in a set of six matrices as described in the following table.

| Matrix      | rows x columns     | Name                           | Description                                              |
|:------------|:-------------------|:-------------------------------|:---------------------------------------------------------|
| **R**       | industry x product | Resource matrix                | Contains exogeneous energy inputs to an ECC              |
| **U**       | product x industry | Use matrix                     | Describes how each energy conversion device uses energy  |
| **U_feed**  | product x industry | Feedstock use matrix           | Describes feedstock inputs to energy conversion devices  |
| **U_EIOU**  | product x industry | Energy industry own use matrix | Describes how the energy industry uses energy            |
| **V**       | industry x product | Make matrix                    | Describes how each energy conversion device makes energy |
| **Y**       | product x industry | Final demand matrix            | Describes how each energy carrier is consumed            |

The matrix approach to data storage brings the challenge that 
different countries, 
and different years for a given country,
have varying energy energy carriers (products) and
varying energy conversion machines (industries).
To get around this challenge, a new `R` package 
[matsbyname](https://matthewheun.github.io/matsbyname/)
enables matrix mathematics that respects row and column names, 
inserting `0` rows or columns where needed.

Furthermore, 
it would be convenient to perform matrix mathematics in `R` data frames
as easily as scalar mathematics in data frames with [tidyverse](https://www.tidyverse.org) syntax. 
[matsindf](https://matthewheun.github.io/matsindf/) enables this functionality.

| Package | Function |
|:--------|:---------|
| [RCLabels](https://matthewheun.github.io/RCLabels/)      | Manipulates row and column names in [matsindf](https://matthewheun.github.io/matsindf/) data frames |
| [matsbyname](https://matthewheun.github.io/matsbyname/)  | Performs matrix mathematics that respects row and column names |
| [matsindf](https://matthewheun.github.io/matsindf/)      | Stores matrices in cells of a data frame, thereby enabling [tidyverse](https://www.tidyverse.org) syntax |

The packages in the table above are applicable to many problem domains and
are available on [CRAN](https://cran.r-project.org).

Additional packages specific to ECC analysis within SEA are available on GitHub
and described in the following table.

| Package | Function |
|:--------|:---------|
| [IEATools](https://matthewheun.github.io/IEATools/) | Converts IEA data to PSUT format |
| [Recca](https://matthewheun.github.io/Recca/)       | Performs `R` energy conversion chain analysis |
| [PFUDatabase](https://energyeconomydecoupling.github.io/PFUDatabase/) | A [targets](https://docs.ropensci.org/targets/) pipeline to create a data frame of PSUT matrices |
| [PFUAggDatabase](https://energyeconomydecoupling.github.io/PFUAggDatabase/) | A [targets](https://docs.ropensci.org/targets/) pipeline to aggregate PSUT matrices |

>>>>

The most important decision for the design of SEA packages
involves data format.
We selected the Physical Supply-Use Table (PSUT) framework,
a matrix approach to describing energy flows from 
resource extraction to
processing stages and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018].
The PSUT framework succinctly describes the flow of energy carriers
("products" in PSUT terminology) 
among energy conversion machines 
("industries" in PSUT terminology)
with a set of six matrices as described in the following table.

| Matrix      | rows x columns     | Name                           | Description                                              |
|:------------|:-------------------|:-------------------------------|:---------------------------------------------------------|
| **R**       | industry x product | Resource matrix                | Contains exogeneous energy inputs to an ECC              |
| **U**       | product x industry | Use matrix                     | Describes how each energy conversion device uses energy; the sum of **U_feed** and **U_EIOU** |
| **U_feed**  | product x industry | Feedstock use matrix           | Describes feedstock inputs to energy conversion devices  |
| **U_EIOU**  | product x industry | Energy industry own use matrix | Describes how the energy industry uses energy            |
| **V**       | industry x product | Make matrix                    | Describes how each energy conversion device makes energy |
| **Y**       | product x industry | Final demand matrix            | Describes how each energy carrier is consumed            |

Further development followed from selection of the PSUT data format.
First, the PSUT framework (a matrix-based data format) brought the challenge that 
different countries and years 
have varying energy energy carriers (products) and
varying energy conversion machines (industries).
I.e., matrices for different countries and years have differing row and column names.
To get around this challenge, we created an `R` package 
([matsbyname](https://github.com/MatthewHeun/matsbyname/)) which
enables matrix mathematics that respects matrix row and column names, 
inserting `0` rows or columns where needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as  _scalar_ mathematics
in `R` data frames
using the [tidyverse](https://www.tidyverse.org) syntax. 
We developed the [matsindf](https://matthewheun.github.io/matsindf/) package 
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels](https://matthewheun.github.io/RCLabels/) package
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN](https://cran.r-project.org).

| Package | Function |
|:--------|:---------|
| [RCLabels](https://github.com/MatthewHeun/RCLabels/)      | Manipulates row and column names in [matsindf](https://github.com/MatthewHeun/matsindf/) data frames |
| [matsbyname](https://github.com/MatthewHeun/matsbyname/)  | Performs matrix mathematics that respects row and column names |
| [matsindf](https://github.com/MatthewHeun/matsindf/)      | Stores matrices in cells of a data frame, thereby enabling [tidyverse](https://www.tidyverse.org) syntax |

Several calculation steps are required to create the PFU database. 
Each step is assisted by an `R` package that we created.
First, the IEA's primary- and final-stage WEEB data must be converted
to the PSUT format, 
a task completed by the [IEATools](https://github.com/MatthewHeun/IEATools/) package. 
Second, human and animal muscle work must be calculated from 
labor and FAO data following the methodology of @Steenwyk:2022ww
using the [MWTools](https://github.com/EnergyEconomyDecoupling/MWTools) package. 
Third, the IEA's primary- and final-stage ECC data 
are extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by the
[PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/) 
and 
[Recca](https://github.com/MatthewHeun/Recca) 
packages.
Finally, ECCs must be converted from energy terms to exergy terms. 
This step is assisted by the [Recca](https://github.com/MatthewHeun/Recca) package.
The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets](https://docs.ropensci.org/targets/)
computation pipeline
available in the [PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/)
package.
These packages are specific to creating the PFU database 
and are summarized in the table below.

| Package | Function |
|:--------|:---------|
| [IEATools](https://github.com/MatthewHeun/IEATools/) | Converts IEA data to [matsindf](https://github.com/MatthewHeun/matsindf/) format |
| [MWTools](https://github.com/energyeconomydecoupling/MWTools) | Converts ILO and FAO data to human and animal muscle work information in [matsindf](https://github.com/MatthewHeun/matsindf/) format |
| [Recca](https://matthewheun.github.io/Recca/)       | Performs `R` energy conversion chain analysis |
| [PFUDatabase](https://energyeconomydecoupling.github.io/PFUDatabase/) | A [targets](https://docs.ropensci.org/targets/) pipeline to create a data frame of PSUT matrices |
| [PFUAggDatabase](https://energyeconomydecoupling.github.io/PFUAggDatabase/) | A [targets](https://docs.ropensci.org/targets/) pipeline to aggregate PSUT matrices |
